---
title: "Lab 10 : R Shiny apps using the COVID-19 reporting data"
output:
  html_document:
    toc: true
    toc_depth : 4
    theme: united
    highlight: tango
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

*I recognize, and fully understand, that this data maybe emotionally difficult to work. My intention is to make these lab relevant, allowing you to gather your own insights directly from new visualizations of the data. Please let me know if you would rather not work with the data.*

## Learning Objectives

* The basics of making Rshiny interactive applications
* Creating a Rshiny app with the covid reporting data

## Shiny Overview and tutorial

The first step is to install the __shiny__ package

You can run an example shiny app by typing the following in the R console

```{r, eval = FALSE}
library(shiny)
runExample("01_hello")
```

There is a great starter set of [7 Shiny tutorials](https://shiny.rstudio.com/tutorial/written-tutorial/lesson1/) written by RStudio that include examples and code snippets related to visualizing the COVID-19 reporting data. I highly recommend going through these.

### Structure of an app.R file

Then let's examing the example new file (which is also the same as 01_hello. Go to File > New File > Shiny Web App

Click on the Run button and the App will be built

app.R has three components:

    a user interface object

    a server function

    a call to the shinyApp function

It you open a new Shiny file you can see these three components

```{r, eval=FALSE}
# Define UI for application that draws a histogram
ui <- fluidPage(...

# Define server logic required to draw a histogram
server <- function(input, output) {...

# Run the application 
shinyApp(ui = ui, server = server)
```

The user interface (ui) object controls the layout and appearance of your app. The server function contains the instructions that your computer needs to build your app. Finally the shinyApp function creates Shiny app objects from an explicit UI/server pair.

Notice the call to load the libraries (in this case __shiny__) comes before the code for the shiny app. Other code that feeds into your app (e.g. loading and manipulating data) also comes before the shiny app.

## Shiny examples using the COVID-19 Reporting data

In the [class GitHub repository](https://github.com/jeffreyblanchard/HumGen2022R) there is folder ShinyApps with examples of shiny apps using the COV-19 `Lab6x_example_5countries_app.R` and `Lab6x_example_map2_app.R` which you can down load to your computer. I have also put these examples on the RStudio Shiny server [Lab10_example_5countries](https://jeffreyblanchard.shinyapps.io/example5countries/) and [Lab10_example_map2](http://jeffreyblanchard.shinyapps.io/ExampleMap). The map example may take a while to load depending on your connections.

## Exercises

Go through the [Shiny tutorials](https://shiny.rstudio.com/tutorial/written-tutorial/lesson1/). Turn your lesson 5 Shiny app (Census Visualization) into a web page using shinyapps.io, RStudio’s hosting service for Shiny apps.  In Moodle send me a link you your app.

